Machine learning methods in streamflow prediction considering climate variable windowing

### 2021-06-28 paper/paper.tex
Adoção das correções feitas no overleaf

\begin{abstract}
Hydropower operation and planning require streamflow forecasts. Machine learning methods combined with atmospheric climate variables complement physical models, particularly, where in-situ observations are not available. To make predictions of streamflow of a reservoir, we test four ML techniques: Support Vector Regressor (SVR), Simple Decision Tree, Random Forest (RF), and Stacking Regressor, which combines SVR, RF, and Ridge. Also, we add prediction horizon varying from 0 to 29 days for streamflow, and windowing of 1, 15, and 30 days for atmospheric climate variables. Results indicate a better fit in the RF model. Models outperform when a shorter horizon prediction of streamflow and longer windowing for climate variables are considered.

@@

In this work, we make short-term forecast streamflow using Três Marias reservoir as study case. We use atmospheric variables as precipitation, evaporation, temperature, and surface runoff, available in Projeta project \cite{inpe}, and observed data of streamflow and water level of Três Marias reservoir \cite{onsnivel,onsvazao}. We compared four machine learning methods in order to access the best performance of short-term predictions. Moreover, we evaluate the impact of adding a horizon prediction of streamflow, as well as, windowing the climate variables.

@@

According to \cite{dou}, ensemble machine learning techniques
provide an accurate and efficient solution in spatial modeling. Stacking ensemble ML algorithm learns how to combine the predictors from other well-performing models. In our case, we combined both SVR, RF models and Ridge regressors.

The natural complexity of hydrologic variables drives the approaches of mathematical models to actual output, in this case, streamflow. Although the selection of input variables varies according to the study, precipitation and temperature are common variables as observed in \cite{kassem2020predicting,essenfelder2020smart,bhatta2019evaluation}. Historical data of streamflow itself as input variable was present in \cite{li}. In order to approach the natural structure of a reservoir, we included in our work the variables runoff and evaporation, besides precipitation, evaporation, streamflow and water level.

@@

We tested four models: SVR, simple decision tree, RF and stacking ensemble method. We evaluated the performance of each model adding a windowing for climate variables in intervals of 1, 15, and 30 days. Furthermore, we tested horizon predictions varying from 0 to 29 days.

@@

We used a combination of parameters to tune the model with an R-squared score as a metric, leveraging the GridSearchCV scikit-learn tool. In GridSearchCV, we define some constraints of each parameter, available in Appendix B. We adopted a cross-validation strategy to optimize these parameters and minimize the risk of overfitting. This was made to avoid testing regressor parameters on the same data it was trained for. The adopted approach was k-fold cross-validation, where the data set was divided into k groups - $k$ was set to 10 in this study - the regressors were trained in $k-1$ of these groups, and the fold left was used to test. As a consequence, the test was made using data the regressor have never seen while training. We calculate the final score as the average of the scores obtained for each test group.

@@

All variables, except month variable, were normalized when evaluating regressors that are sensitive to data scale, in order to prevent them from being dominated by variables with higher order of magnitude. We used the Min-Max normalization given by the Equation
    \begin{equation}
    \label{eqn:Normalization}
    x' = \frac{x-min(x)}{max(x)-min(x)}
    \end{equation}
to adjust the scale of the inputs. Where the value $x$ is the present value, $min(x)$ is the minimum value of the feature, and $max(x)$ is the maximum value of the feature. Then, we transformed that to their original scale.

@@

Having tested horizon prediction of streamflow and water level varying from 1 to 29 days, and windowing at intervals of 1, 15 and 30 days for the climate variables, results indicated better fit for horizon of one day and window of 30 days. However, the models presented different results of R\textsuperscript{2} when they are compared to each other, as shown in Figure~\ref{fig:r2}.

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/R2_w01}
        \captionsetup{margin={.15\textwidth, 0cm}}
        \caption{1 day windowing}
        \label{fig:r2_1}
    \end{subfigure}
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/R2_w15}
        \captionsetup{margin={.15\textwidth, 0cm}}
        \caption{15 days windowing}
        \label{fig:r2_15}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includesvg{Graphs/R2_w30}
        \captionsetup{justification=justified,singlelinecheck=false}
        \captionsetup{margin={.23\linewidth, 0cm}}
        \caption{30 days windowing}
        \label{fig:r2_30}
    \end{subfigure}
    \caption{R\textsuperscript{2} for \subref{fig:r2_1} 1, \subref{fig:r2_15} 15, and \subref{fig:r2_30} 30 days windowing}
    \label{fig:r2}
\end{figure}

According to the increase of prediction horizon of streamflow up to 8 days, the R\textsuperscript{2} had an accentuated decrease in all models in 1 day-windowing. This trend is kept in 15 and 30 days windowing only for Decision Tree, while for SVR and Stacking Regressor the decrease is smaller and for Random Forest, R\textsuperscript{2} presented a growth.

@@

From an evaluation of Figure~\ref{fig:StackedComparisson}, which demonstrates the true versus predicted values for 1 and 29 days of prediction horizon, it is evident that the model trained for 1 day of horizon presents greater sensitivity to sudden changes in reservoir streamflow, while that the model with the longest horizon time, despite being able to follow the streamflow trend, does not faithfully represent the peaks present. Another point that can be justified by the presence of these peaks is the relatively large value found for the MAE since the model ends up being highly penalized by sudden changes in streamflow. On the other hand, an R-squared metric, which considers the amplitude of the actual data, ends up not being penalized in the same way. Figure~\ref{fig:StackedComparisson} presents data for a two-year window from 2000 to 2002, so data can be more easily visualized. For a visualization for the whole data range, see Appendix~\ref{sec:TrueVersusPredicted}

@@

After 8 days of prediction horizon, the R\textsuperscript{2} tends to be constant in 15 and 30 days-windowing cases. In 1 day-windowing case, the models presented an unexpected performance, since the models had a turning point in 15 days-horizon, in which the performance reached its lowest point and rose again.

Random forest outperforms in all considered cases of prediction horizon and windowing in terms of R\textsuperscript{2}. Decision Tree Regressor has the worst fit, while SVR and Stacking Regressor had similar results, although Stacking Regressor slightly presents higher values of R\textsuperscript{2}.

@@

In general terms, RF had a better results of MAE, if compared to other models. SVR presented the worst result, while both Stacking Regressor and Decision Tree had similar results.

@@

In this work, we tested four methods to forecast streamflow using Três Marias reservoir as a study case. Results indicated that RF had the best fit over the models.  For one-day prediction horizon for streamflow and water level , models presented R-squared values in the range 0.7 to 0.9, but that decreases according to the increase of  prediction horizon. With regards to windowing of climate variables, the best fits occur when considering a 30 days-windowing, which might be related to the delay of the physical process impact of all climate variables in the entire sub-basin on the reservoir upstream.

SVR had the highest values of MAE, while RF had the lowest values. The Stacking Regressor and decision tree produced similar values of MAE. Surprisingly, RF presented a slight increase of MAE for increased prediction horizon, and for the case of 15 and 30 days-windowing, while the other models presented a trend of having a decline.

Empirical results obtained from these models showed the performance not only over the models, but also changing the lag of streamflow and water level, and the window period of climate variables. That is important in solving water resources management problems and improve the streamflow prediction.

>>>>

\begin{abstract}
Hydropower operation and planning require streamflow forecasts. Machine learning (ML) methods combined with atmospheric climate variables complement physical models, particularly, where in-situ observations are not available. To make predictions of streamflow of a reservoir, we test four ML techniques: Support Vector Regressor (SVR), Simple Decision Tree, Random Forest (RF), and Stacking Regressor, which combines SVR, RF, and Ridge. Also, we add prediction horizon varying from 1 to 29 days for streamflow, and windowing of 1, 15, and 30 days for atmospheric climate variables. Results indicate a better fit in the RF model. Models outperform when a shorter horizon prediction of streamflow and longer windowing for climate variables are considered.

@@

In this work, we explore machine learning methods for short-term forecast streamflow using Três Marias reservoir as a case study. The objective here is to evaluate the performance of Support Vector Regressor (SVR), Simple Decision Tree, Random Forest (RF), and Stacking Regressor, which combines SVR, RF, and Ridge, in different prediction horizons and climate variable windowing. We use atmospheric variables as precipitation, evaporation, temperature, and surface runoff, available in Projeta project \cite{inpe}, and observed data of streamflow and water level of Três Marias reservoir \cite{onsnivel,onsvazao}.

@@

According to \cite{dou}, ensemble machine learning techniques
provide an accurate and efficient solution in spatial modeling. Stacking ensemble ML algorithm learns how to combine the predictors from other well-performing models. In our case, we combine both SVR, RF models and Ridge regressors.

The natural complexity of hydrologic variables drives the approaches of mathematical models to actual output, in this case, streamflow. Although the selection of input variables varies according to the study, precipitation and temperature are common variables as observed in \cite{kassem2020predicting,essenfelder2020smart,bhatta2019evaluation}. Historical data of streamflow itself as input variable was present in \cite{li}. In order to approach the natural structure of a reservoir, we include in our work the variables runoff and evaporation, besides precipitation, evaporation, streamflow and water level.

@@

We tested four models: SVR, simple decision tree, RF and stacking ensemble method. We evaluated the performance of each model adding a windowing for climate variables in intervals of 1, 15, and 30 days. Furthermore, we tested horizon predictions varying from 1 to 29 days.

@@

We used a combination of parameters to tune the model with an R-squared score as a metric, leveraging the GridSearchCV scikit-learn tool. In GridSearchCV, we define some constraints of each parameter, detailed in Appendix B. We adopted a cross-validation strategy to optimize these parameters and minimize the risk of overfitting. This was made to avoid testing regressor parameters on the same data it was trained for. The adopted approach was K-Fold cross-validation, where the data set was divided into k groups - $k$ was set to 10 in this study - the regressors were trained in $k-1$ of these groups, and the fold left was used to test. As a consequence, the test data used by the regressors were new for every new training session. We calculate the final score as the average of the scores obtained for each test group.

@@

All variables, except month, were normalized when evaluating regressors that are sensitive to data scale, in order to prevent them from being dominated by variables with higher order of magnitude. We used the Min-Max normalization given by the Equation
    \begin{equation}
    \label{eqn:Normalization}
    x' = \frac{x-min(x)}{max(x)-min(x)}
    \end{equation}
to adjust the scale of the inputs. Where the value $x$ is the present value, $min(x)$ is the minimum value of the feature, and $max(x)$ is the maximum value of the feature. Then, we transformed that to their original scale.

@@

Having tested horizon prediction of streamflow and water level varying from 1 to 29 days, and windowing at intervals of 1, 15 and 30 days for the climate variables, results indicated better fit for horizon of one day and window of 30 days. However, the models presented different results of R-squared when they are compared to each other, as shown in Figure~\ref{fig:r2}.

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/R2_w01}
        \captionsetup{margin={.15\textwidth, 0cm}}
        \caption{1 day windowing}
        \label{fig:r2_1}
    \end{subfigure}
    \begin{subfigure}[b]{.49\textwidth}
        \centering
        \includesvg{Graphs/R2_w15}
        \captionsetup{margin={.15\textwidth, 0cm}}
        \caption{15 days windowing}
        \label{fig:r2_15}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includesvg{Graphs/R2_w30}
        \captionsetup{justification=justified,singlelinecheck=false}
        \captionsetup{margin={.23\linewidth, 0cm}}
        \caption{30 days windowing}
        \label{fig:r2_30}
    \end{subfigure}
    \caption{R-squared for \subref{fig:r2_1} 1, \subref{fig:r2_15} 15, and \subref{fig:r2_30} 30 days windowing}
    \label{fig:r2}
\end{figure}

According to the increase of prediction horizon of streamflow up to 8 days, the R-squared had an accentuated decrease in all models in 1 day-windowing. This trend is kept in 15 and 30 days windowing only for Decision Tree, while for SVR and Stacking Regressor the decrease is smaller and for Random Forest, R-squared presented a growth.

@@

From an evaluation of Figure~\ref{fig:StackedComparisson}, which demonstrates the true versus predicted values for 1 and 29 days of prediction horizon, it is evident that the model trained for 1 day of horizon presents greater sensitivity to sudden changes in reservoir streamflow, while that the model with the longest horizon time, despite being able to follow the streamflow trend, does not faithfully represent the peaks present. Another point that can be justified by the presence of these peaks is the relatively large value found for the MAE since the model ends up being highly penalized by sudden changes in streamflow. On the other hand, an R-squared metric, which considers the amplitude of the actual data, ends up not being penalized in the same way. Figure~\ref{fig:StackedComparisson} presents data for a two-year window from 2000 to 2002, so it can be more easily visualized. For a visualization for the whole data range, see Appendix~\ref{sec:TrueVersusPredicted}

@@

After 8 days of prediction horizon, the R-squared tends to be constant in 15 and 30 days-windowing cases. In 1 day-windowing case, the models presented an unexpected performance, since the models had a turning point in 15 days-horizon, in which the performance reached its lowest point and rose again.

Random forest outperforms in all considered cases of prediction horizon and windowing in terms of R-squared. Decision Tree Regressor has the worst fit, while SVR and Stacking Regressor had similar results, although Stacking Regressor slightly presents higher values of R-squared.

@@

In general terms, RF had a better results of MAE and R-squared, if compared to other models. SVR presented the worst result, while both Stacking Regressor and Decision Tree had similar results.

@@

In this work, we tested four methods to forecast streamflow using Três Marias reservoir as a case study. Empirical results obtained from these models showed the performance not only over the models, but also changing the prediction horizon of streamflow, and the window period of climate variables. That is important in solving water resources management problems and improve streamflow prediction.

Results indicated that RF had the best fit over the other models. For one-day prediction horizon for streamflow, models presented R-squared values in the range 0.7 to 0.9, but that decreases according to the increase of  prediction horizon. With regards to windowing of climate variables, the best fits occur when considering a 30 days-windowing, which might be related to the delay of the physical process impact of all climate variables in the entire sub-basin on the reservoir upstream.

SVR had the highest values of MAE, while RF had the lowest values. The Stacking Regressor and decision tree produced the 2nd and 3rd best MAE results, respectively. Surprisingly, RF presented a slight increase of MAE for increased prediction horizon, and for the case of 15 and 30 days-windowing, while the other models presented a trend of having a decline.
